Adaptive evolution with aneuploidy and mutation

### 2022-01-09 ms/ms.tex
updated 2n* fixation plot; added notebooks; updated estimated MAPs, because default tau=1

\paragraph{Parameter estimation.} 
We used ABC-SMC to infer the posterior distribution of the parameters of the single-locus model.
The MAP (maximum a posteriori) and 50\% HDI (highest density interval) of the parameters posterior are~(\autoref{fig:posterior}):
mutation rate, $\mu=3.06\substack{+1.229 \\ -2.84}\times10^{-6}$,
aneuploidy rate, $\delta=1.61\substack{+1.122 \\ -0.197}\times10^{-3}$,
and fitness values, $w_{2n+1}=1.022\substack{+0.001 \\ -0.002}$,
$w_{2n+1^*}=1.025\substack{+0.001 \\ -0.001}$,
$w_{2n^*}=1.028\substack{+0.001 \\ -0.002}$, all relative to the fitness of \euwt, which is set to $w_{2n}=1$.
% describe b and c
This estimated aneuploidy rate agrees with previous estimates. % TODO refs
The mutation rate corresponds to a mutation target size of $10^{4}$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$~\citep{Zhu2014}. 
%We can see from the posterior distribution~(\autoref{fig:posterior}) the correlation between fitness of different genotypes. 
% TODO say/plot something about the correlation: which fitness values are correlated? how strong is it? why does it occur?

We also tried to inference parameters of the single-locus model with different $\tau$: 1, 2, 5, 10 and 100. It was seen that for 1, 2, 5 the posterior was not significantly different from the model with $\tau=33/32$, while for $\tau=10$ the estimated $\mu\ $ was a little lower, $1.654\substack{+0.879 \\ -1.775}\times10^{-6}$, and for $\tau=100$ the estimated $\mu\ $ was already significantly lower and $\delta\ $ had higher variability. The MAP and 50\% HDI for $\tau=100$:
$\mu=5.538\substack{+1.611 \\ -5.311}\times10^{-7}$,
$\delta=8.292\substack{+8.988 \\ -4.96}\times10^{-4}$,
$w_{2n+1}=1.023\substack{+0.001 \\ -0.002}$,
$w_{2n+1^*}=1.024\substack{+0.001 \\ -0.001}$,
$w_{2n^*}=1.028\substack{+0.001 \\ -0.002}$.
%TODO: show joint distributions comparison plot (20-diffk.ipynb). WAIC?

@@

However, a model without aneuploidy where the aneuploidy rate is fixed at zero, $\delta=0$, cannot explain the experimental observations~(Figures~\ref{fig:2n*-fixation} and \ref{fig:likelihood}). The estimated parameter values without aneuploidy are:
mutation rate, $\mu=1.251\substack{+0.146 \\ -0.177}\times10^{-8}$,
fitness of mutant, $w_{2n^*}=1.011\substack{+0.001 \\ -0.001}$. 
We can see that the mutation rate estimate in this model is much lower than in the model with aneuploidy. Higher mutation rate would cause quicker fixation of \eumt\ than in the experimental observations (\autoref{fig:2n*-fixation}). Thus, even disregarding the appearance of trisomic cells in the experiment, the evidence supports the inclusion of aneuploidy in the model.
% TODO WAIC?

@@

  \caption{
  % TODO change x label - "Time to fixation of 2n*"
  % plot both lines on one panel with two colors (blue and red for with and without? choose nice color, not the default 'r' and 'b'!), make the without in dashed.
  % place 82% a bit higher
  % xticks - replace 500, 1000,1500,2000,2500 with 450, 1700, 2350
    \textbf{Single-locus model fit with and without aneuploidy.} Cumulative distribution of the time to fixation of \eumt\ in 10,000 simulations using the MAP estimate with and without aneuploidy (i.e., $\delta=0$ is fixed).
     The MAP likelihood (\autoref{eq:heatstress-likelihood}) is 0.86 and 0.75 with and without aneuploidy.
	}
  \label{fig:likelihood}
\end{figure}

@@

\begin{figure}[h!]
  \centering
\includegraphics[width=0.6\textwidth]{../figures/fixation-plot.pdf}
  \caption{
  \textbf{Adaptation time cannot be explained without aneuploidy.}
The distribution of adaptation time (fixation of \eumt) in $10,000$ simulations of the single-locus model with aneuploidy (blue; MAP parameters) compared to two models without aneuploidy: a model with the same parameter values except $\delta=0$ (orange); and a model fitted to the data assuming $\delta=0$ (green).
In the experiment by \citet{Yona2012}, one population lost aneuploidy by generation 1,700 and the other by generation 2,350 (dashed lines) but not before generation 200. Thus, the blue distribution demonstrates a better fit compared to the green, and the yellow demonstrates a very poor fit.
The last bin contains all the simulations with time equal or greater than 3,000.
}
% TODO add legend
% TODO change y label to % simulations
% change x label to Time to fixation of 2n*
% makes vertical line black
% add line for t=200?
% does x axis start at zero? can you add 0 to x ticks?
  \label{fig:2n*-fixation}
\end{figure}

@@

% Fig ABC iterations
\begin{figure}[h]

>>>>

\paragraph{Parameter estimation.} 
We used ABC-SMC to infer the posterior distribution of the parameters of the single-locus model.
The MAP (maximum a posteriori) and 50\% HDI (highest density interval) of the parameters posterior are~(\autoref{fig:posterior}): %% generated by MAP-and-hdi.ipynb
mutation rate, $\mu=2.93\substack{+1.122 \\ -2.824}\times10^{-6}$,
aneuploidy rate, $\delta=1.716\substack{+0.974 \\ -0.384}\times10^{-3}$,
and fitness values, $w_{2n+1}=1.022\substack{+0.001 \\ -0.002}$,
$w_{2n+1^*}=1.025\substack{+0.001 \\ -0.001}$,
$w_{2n^*}=1.028\substack{+0.001 \\ -0.001}$, all relative to the fitness of \euwt, which is set to $w_{2n}=1$.
% describe b and c TODO? what does it mean?
This estimated aneuploidy rate agrees with previous estimates. % TODO refs
The mutation rate corresponds to a mutation target size of $10^{4}$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$~\citep{Zhu2014}. 
%We can see from the posterior distribution~(\autoref{fig:posterior}) the correlation between fitness of different genotypes. 
% TODO say/plot something about the correlation: which fitness values are correlated? how strong is it? why does it occur?

We also tried to inference parameters of the single-locus model with different $\tau$: 1, 2, 5, 10 and 100. It was seen that for 1, 2, 5 the posterior was not significantly different from the model with $\tau=33/32$, while for $\tau=10$ the estimated $\mu\ $ was a little lower, $1.654\substack{+0.879 \\ -1.775}\times10^{-6}$, and for $\tau=100$ the estimated $\mu\ $ was already significantly lower and $\delta\ $ had higher variability. The MAP and 50\% HDI for $\tau=100$: %TODO add to MAP-and-hdi.ipynb and check the values here
$\mu=5.538\substack{+1.611 \\ -5.311}\times10^{-7}$,
$\delta=8.292\substack{+8.988 \\ -4.96}\times10^{-4}$,
$w_{2n+1}=1.023\substack{+0.001 \\ -0.002}$,
$w_{2n+1^*}=1.024\substack{+0.001 \\ -0.001}$,
$w_{2n^*}=1.028\substack{+0.001 \\ -0.002}$.
%TODO: show joint distributions comparison plot (20-diffk.ipynb). WAIC?

@@

However, a model without aneuploidy where the aneuploidy rate is fixed at zero, $\delta=0$, cannot explain the experimental observations~(Figures~\ref{fig:2n*-fixation} and \ref{fig:likelihood}). The estimated parameter values without aneuploidy are: %% generated by MAP-and-hdi.ipynb
mutation rate, $\mu=7.837\substack{+0.078 \\ -0.123}\times10^{-9}$,
fitness of mutant, $w_{2n^*}=1.013\substack{+0.000 \\ -0.000}$. 
We can see that the mutation rate estimate in this model is much lower than in the model with aneuploidy. Higher mutation rate would cause quicker fixation of \eumt\ than in the experimental observations (\autoref{fig:2n*-fixation}). Thus, even disregarding the appearance of trisomic cells in the experiment, the evidence supports the inclusion of aneuploidy in the model.
% TODO WAIC?

@@

  \caption{
  % TODO change x label - "Time to fixation of 2n*"
  % plot both lines on one panel with two colors (blue and red for with and without? choose nice color, not the default 'r' and 'b'!), make the without in dashed.
  % place 82% a bit higher
  % xticks - replace 500, 1000,1500,2000,2500 with 450, 1700, 2350
    \textbf{Single-locus model fit with and without aneuploidy.} Cumulative distribution of the time to fixation of \eumt\ in 10,000 simulations using the MAP estimate with and without aneuploidy (i.e., $\delta=0$ is fixed).
     The MAP likelihood (\autoref{eq:heatstress-likelihood}) is 0.84 and 0.67 with and without aneuploidy.
	}
  \label{fig:likelihood}
\end{figure}

@@

%%% generated by with-aneuploidy_vs_no_aneuploidy.ipynb
\begin{figure}[h!]
  \centering
\includegraphics[width=0.9\textwidth]{../figures/fixation-plot.pdf}
  \caption{
  \textbf{Adaptation time cannot be explained without aneuploidy.}
The distribution of adaptation time (fixation of \eumt) in $10,000$ simulations of the single-locus model with aneuploidy (blue; MAP parameters) compared to two models without aneuploidy: a model with the same parameter values except $\delta=0$ (orange); and a model fitted to the data assuming $\delta=0$ (green).
In the experiment by \citet{Yona2012}, one population lost aneuploidy by generation 1,700 and the other by generation 2,350 (dashed lines) but not before generation 200. Thus, the blue distribution demonstrates a better fit compared to the green, and the yellow demonstrates a very poor fit.
The last bin contains all the simulations with time equal or greater than 3,000.
}
  \label{fig:2n*-fixation}
\end{figure}

@@

% Fig ABC iterations, convergence
%% generated by convergence.ipynb
\begin{figure}[h]
